Multilevel generalizations of collective spin dynamics and squeezing

### 2019-11-03 collective_sun/multilevel_spin_notes.tex
simplified derivation of effective SOC hamiltonian

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Z}{\mathcal{Z}}

@@

In addition to the fact that the fully symmetric manifold $\M_\FS$ is
ground-state manifold of the interaction Hamltonian $H_{\t{int}}$, an
important feature of $\M_\FS$ is its closure under the action of
unitaries generated by collective operators of the form
$\Q=\sum_pQ^{(p)}$, as well as linear combinations and products
thereof.  Closure of $\M_\FS$ under collective dynamics is a
straightforward consequence of the permutational symmetry obeyed by
collective operators.

@@

The single-particle Hamiltonian $H_{\t{lat}}^{(\phi)}$ breaks
permutational symmetry, thereby coupling states within the fully
symmetric manifold $\M_\FS$ to states outsides it.  If SOC is
sufficiently weak, however, transitions outside the fully symmetric
manifold $\M_\FS$ are energetically suppressed by interactions, and we
can account for the effect of SOC perturbatively.  In the regime of
weak SOC with $\abs{\phi}\ll1$, by considering
\begin{enumerate*}
\item all $L\le12$ for arbitrary $n$, and
\item all $L<n$ for all $n\le25$
\end{enumerate*}
(and conjecturing the same functional form for all $L$ and $n$), we
find that
% expand Clebsch-Gordan coefficients in a power series according to
% \url{functions.wolfram.com/07.38.06.0008.01}
\begin{align}
  A_L^{(\phi)}
  = \p{-1}^{L+\lfloor L/2\rfloor} \xi_L \phi^L + O\p{\phi^{L+2}},
  \label{eq:A_L_phi_small}
\end{align}
where $\xi_L$ is a scale factor defined in \eqref{eq:scale_fac}.  A
perturbative treatment of SOC through second order in the SOC angle
$\phi$ yields the following effective Hamiltonian in the fully
symmetric manifold $\M_\FS$ (see Appendix \ref{sec:SOC_pert}):
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
\end{align}
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_q\sp{\sin\p{qa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi}
\end{align}
where
\begin{align}
  \EE_q\sp{X_q} \equiv \f1N \sum_q X_q,
  &&
  \var_q\sp{X_q} \equiv \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}}
  = \EE_p\sp{X_p^2}-\EE_q\sp{X_q}^2,
\end{align}
are the mean and variance of $X_q$ over all occupied quasi-momenta
$q$.  In the case of SU(2), the effective Hamiltonians in
\eqref{eq:H_eff_1_phi} and \eqref{eq:H_eff_2_phi} reduce to the
spin-squeezing Hamiltonians derived in Ref.~\cite{he2019engineering}.
To facilitate comparison, we note that in the case of SU(2) the drive
operator $\D_{1,0}$ is precisely the collective spin-$z$ operator
$S_\z$ that appears in Ref.~\cite{he2019engineering}.  Furthermore,
drive operators $D_{LM}$ vanish when $L\ge n$, so in the case of SU(2)
the operator $\D_{2,0}$ has no contribution to the effective
Hamiltonians in \eqref{eq:H_eff_2_phi} and \eqref{eq:H_eff_2_phi}.

@@

Here we derive the effective Hamiltonian induced on the fully
symmetric manifold $\M_\FS$ by the single-particle SOC Hamiltonian
$H_{\t{lat}}^{(\phi)}$ in \eqref{eq:H_lat_SOC_B} in the weak SOC limit
$\abs{\phi}\ll1$.  For brevity, we work in units with the lattice
spacing $a=1$ throughout this section.  Perturbations $V$ to the
interaction Hamiltonian $H_{\t{int}}$ induce the following effective
Hamiltonians on the fully symmetric manifold
$\M_\FS$\cite{bravyi2011schrieffer, perlin2019effective}:
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 V \P_0,
  &&
  H_{\t{eff}}^{(2)} = -\P_0 V \E V \P_0,
  &&
  \E \equiv \sum_{k>0} \f{\P_k}{\Delta_k},
  \label{eq:pert}
\end{align}
where $\P_k$ is a projector onto the $k$-th excited eigenspace of the
interaction Hamiltonian $H_{\t{int}}$, with interaction energy
$\Delta_k$ above that of fully symmetric states
$\ket{\psi_\FS}\in\M_\FS$.

@@

The single-particle SOC Hamiltonian is
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -t \sum_{q,L} w_L\p{q} A_L^{(\phi)} D_{L,0}^{(q)},
  &&
  w_L\p{q} \equiv
  \begin{cases}
    \cos q & L~\t{even} \\
    \sin q & L~\t{odd}
  \end{cases},
  \label{eq:H_lat_SOC_A}
\end{align}
where $A_L^{(\phi)}$ is essentially a sum over Clebsch-Gordan
coefficients, defined in \eqref{eq:A_L_phi}, and scales as
$A_L^{(\phi)} \sim \phi^L + O\p{\phi^{L+2}}$ when $\abs{\phi}\ll1$.
When the SOC angle $\phi$ is sufficiently small in magnitude, the
effective Hamiltonian induced on the fully symmetric manifold $\M_\FS$
at first order in the perturbation $H_{\t{lat}}^{(\phi)}$ is
\begin{align}
  H_{\t{eff}}^{(1)}
  = \P_0 H_{\t{lat}}^{(\phi)} \P_0
  = -t \sum_{q,L} w_L\p{q} A_L^{(\phi)} \P_0 D_{L,0}^{(q)} \P_0,
\end{align}
where we can freely change the index of each single-mode drive
operator $D_{L,0}^{(q)}$ within the fully symmetric manifold to
re-write
\begin{align}
  \P_0 D_{L,0}^{(q)} \P_0 = \f1N \sum_p \P_0 D_{L,0}^{(p)} \P_0
  = \f1N \P_0 \D_{L,0} \P_0,
\end{align}
which implies that
\begin{align}
  H_{\t{eff}}^{(1)}
  = -t \sum_L \EE_q\sp{w_L\p{q}} A_L^{(\phi)} \D_{L,0},
  &&
  \EE_q\sp{X_q} \equiv \f1N \sum_q X_q.
  \label{eq:H_eff_1}
\end{align}
Here $\EE_q\sp{X_q}$ is an average of $X_q$ over all occupied
quasi-momenta $q$.  Note that we have dropped the explicit projection
operators $\P_0$ in \eqref{eq:H_eff_1} with the understanding that we
are implicitly restricting all effective dynamics to the fully
symmetric manifold $\M_\FS$.

@@

In order to expand the second-order effective Hamiltonian
$H_{\t{eff}}^{(2)}$ induced on the fully symmetric manifold $\M_\FS$
by the SOC Hamiltonian $H_{\t{lat}}^{(\phi)}$ in
\eqref{eq:H_lat_SOC_A} when the SOC angle $\abs{\phi}\ll1$, we note
that $w_L\p{q}$ in \eqref{eq:H_lat_SOC_A} takes the form
\begin{align}
  w_L\p{q} = \f12 \sum_{s\in\set{\pm1}} \chi_{Ls} e^{isq},
  \label{eq:w_chi}
\end{align}
for some appropriately defined coefficients $\chi_{Ls}$, which implies
that
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -\f{t}{2} \sum_{\substack{L\\s\in\set{\pm1}}}
  \chi_{Ls} A_L^{(\phi)} \Z_{Ls},
  &&
  \Z_{Ks} \equiv \sum_q e^{isq} D_{L,0}^{(q)},
\end{align}
and in turn
\begin{align}
  H_{\t{eff}}^{(2)}
  = -\f{t^2}{4} \sum_{\substack{J,K\\r,s\in\set{\pm1}}}
  \chi_{Jr} \chi_{Ks} A_J^{(\phi)} A_K^{(\phi)}
  \P_0 \Z_{Jr} \E \Z_{Ks} \P_0,
  \label{eq:H_eff_2_start}
\end{align}
where $\P_0$ is a projector onto the fully symmetric manifold
$\M_\FS$, and $\E$ is a sum over projectors onto excited-state
manifolds with corresponding energetic suppression factors, defined in
\eqref{eq:pert}.  The remainder of this section is devoted to the task
of simplifying \eqref{eq:H_eff_2_start} to arrive at our final result
in \eqref{eq:H_eff_2}, expressed in purely in terms of collective
drive operators that are implicitly restricted to the fully symmetric
manifold $\M_\FS$.  We will break up our task into three steps:
\begin{enumerate*}
\item eliminating the excited-state projectors in $\E$,
\item evaluating the product of operators in \eqref{eq:H_eff_2_start},
  and
\item evaluating the sum over $r,s\in\set{\pm1}$.
\end{enumerate*}

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Mean-subtracted spin-wave operators and the elimination
  of excited-state projectors}

In order to simplify the product of operators in
\eqref{eq:H_eff_2_start}, we note that sandwiching the ``spin-wave''
operators $\Z_{Jr},\Z_{Ks}$ by the ground-state projector $\P_0$ and
the excited-state projectors in $\E$, picks off the components of
these spin-wave operators that are off-diagonal with respect to the
symmetric manifold $\M_\FS$ and its orthogonal complement.  We
therefore define the mean-subtracted spin-wave operators
  \tilde\Z_{Ks}
  \equiv \sum_q \p{e^{isq} - \EE_k\sp{e^{isk}}} D_{K,0}^{(q)},
  &&
  \EE_k\sp{X_k} \equiv \f1N \sum_k X_k,
for which the difference $\tilde\Z_{Ks}-\Z_{Ks}\propto\D_{K,0}$ is a
collective operator that only couples fully symmetric states
$\ket{\psi_\FS}\in\M_\FS$ to fully symmetric states.  In particular,
  \P_0 \tilde\Z_{Jr} \E \tilde\Z_{Ks} \P_0
  = \P_0 \Z_{Jr} \E \Z_{Ks} \P_0,
which allows us to freely take
$\Z_{Jr},\Z_{Ks}\to\tilde\Z_{Jr},\tilde\Z_{Ks}$ with no consequence
for the effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:H_eff_2_start}.  Mean-subtracted spin-wave operators
$\tilde\Z_{Ks}$ have the convenient feature that their action on any
fully symmetric state $\ket{\psi_{\FS}}\in\M_\FS$ yields an
eigenvector of the interaction Hamiltonian
$H_{\t{int}}=-u\sum_{p<q}\Pi_{pq}$.  To substantiate this claim, we
expand
  \sum_{p<q} \Pi_{pq} \tilde\Z_{Ks} \ket{\psi_\FS}
  = \sum_{\substack{p<q\\k}}
  \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  \Pi_{pq} D_{K,0}^{(k)} \ket{\psi_\FS}.
  \label{eq:Z_eig_start}
If $k\notin\set{p,q}$, then the permutation operator $\Pi_{pq}$
commutes with $D_{K,0}^{(k)}$, and in turn acts trivially on the
permutationally-symmetric state $\ket{\psi_\FS}$, giving us
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  \Pi_{pq} D_{K,0}^{(k)} \ket{\psi_\FS}
  = { N-1 \choose 2 } \tilde\Z_{Ks} \ket{\psi_\FS}.
  \label{eq:Z_eig_notin}
If $k\in\set{p,q}$, then we can resolve the identity operator
$\1=\Pi_{pq}^\dag \Pi_{pq}$ to expand
\begin{multline}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  \Pi_{pq} D_{K,0}^{(k)} \ket{\psi_\FS}
  = \f12 \sum_{\substack{p\ne q\\k\in\set{p,q}}}
  \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  \Pi_{pq} D_{K,0}^{(k)} \Pi_{pq}^\dag \Pi_{pq} \ket{\psi_\FS} \\
  = \sum_q \sum_{k\ne q} \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  D_{K,0}^{(q)} \ket{\psi_\FS},
\end{multline}
where the sum over $k\ne q$ is equal to a sum over all $k$ minus a
term with $k=q$, i.e.
  \sum_{k\ne q} \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  = \sum_k \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  - \p{e^{isq} - \EE_\ell\sp{e^{is\ell}}}
  = - \p{e^{isq} - \EE_\ell\sp{e^{is\ell}}},
so
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  \Pi_{pq} D_{K,0}^{(k)} \ket{\psi_\FS}
  = - \tilde\Z_{Ks} \ket{\psi_\FS}.
  \label{eq:Z_eig_in}
Substituting the results in \eqref{eq:Z_eig_in} and
\eqref{eq:Z_eig_notin} into \eqref{eq:Z_eig_start} gives us
  \sum_{p<q} \Pi_{pq} \tilde\Z_{Ks} \ket{\psi_\FS}
  = \sp{{ N-2 \choose 2 } - 1} \tilde\Z_{Ks} \ket{\psi_\FS}.
The image of any fully symmetric state $\ket{\psi_\FS}\in\M_\FS$ under
the action of the mean-subtracted spin-wave operator $\tilde\Z_{Ks}$
is thus proportional to an eigenstate of $H_{\t{int}}$ with
interaction energy $-u\sp{{ N-1 \choose 2 } - 1}$, which is $Nu$ above
the interaction energy of $-u{N\choose2}$ of states in the fully
symmetric manifold, so
  \E \tilde\Z_{Ks} \P_0
  =  \f1{Nu} \tilde\Z_{Ks} \P_0
  = \f1{fU} \tilde\Z_{Ks} \P_0,
  \label{eq:proj_elim}
where $f$ is the filling fraction of lattice sites.  We have thus far
simplified the effective Hamiltonian in \eqref{eq:H_eff_2_start} to
  H_{\t{eff}}^{(2)}
  = -\f{t^2}{4fU} \sum_{\substack{J,K\\r,s\in\set{\pm1}}}
  \chi_{Jr} \chi_{Ks} A_J^{(\phi)} A_K^{(\phi)}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0,
  \label{eq:H_eff_2_mid}
leaving us the task of evaluating the product
$\P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0$, as well as the sum over
$r,s$.

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{The spin-wave operator product}

>>>>

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Z}{\mathcal{Z}}

@@

In addition to the fact that the fully symmetric manifold $\M_\FS$ is
ground-state manifold of the interaction Hamltonian $H_{\t{int}}$, an
important feature of $\M_\FS$ is its closure under the action of
collective operators of the form $\Q=\sum_pQ^{(p)}$, as well as linear
combinations and products thereof.  Closure of $\M_\FS$ under
collective the action of collective operators, and in particular
collective dynamics, is a straightforward consequence of the
permutational symmetry obeyed by collective operators.

@@

The single-particle Hamiltonian $H_{\t{lat}}^{(\phi)}$ breaks
permutational symmetry, thereby coupling states within the fully
symmetric manifold $\M_\FS$ to states outsides it.  If SOC is
sufficiently weak, however, transitions outside the fully symmetric
manifold $\M_\FS$ are energetically suppressed by interactions, and we
can account for the effect of SOC perturbatively.  In the regime of
weak SOC with $\abs{\phi}\ll1$, by considering
\begin{enumerate*}
\item all $L\le12$ for arbitrary $n$, and
\item all $L<n$ for all $n\le25$
\end{enumerate*}
(and conjecturing the same functional form for all $L$ and $n$), we
find that
% expand Clebsch-Gordan coefficients in a power series according to
% \url{functions.wolfram.com/07.38.06.0008.01}
\begin{align}
  A_L^{(\phi)}
  = \p{-1}^{L+\lfloor L/2\rfloor} \xi_L \phi^L + O\p{\phi^{L+2}},
  \label{eq:A_L_phi_small}
\end{align}
where $\xi_L$ is a scale factor defined in \eqref{eq:scale_fac}.  A
perturbative treatment of SOC through second order in the SOC angle
$\phi$ yields the following effective Hamiltonian in the fully
symmetric manifold $\M_\FS$ (see Appendix \ref{sec:SOC_pert}):
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
\end{align}
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_q\sp{\sin\p{qa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi}
\end{align}
where
\begin{align}
  \EE_q\sp{X_q} \equiv \f1N \sum_q X_q,
  &&
  \var_q\sp{X_q} \equiv \EE_q\sp{\p{X_q-\EE_k\sp{X_k}}}
  = \EE_q\sp{X_q^2}-\EE_k\sp{X_k}^2,
\end{align}
are the mean and variance of $X_q$ over all occupied quasi-momenta
$q$.  In the case of SU(2), the effective Hamiltonians in
\eqref{eq:H_eff_1_phi} and \eqref{eq:H_eff_2_phi} reduce to the
spin-squeezing Hamiltonians derived in Ref.~\cite{he2019engineering}.
To facilitate comparison, we note that in the case of SU(2) the
re-scaled drive operator $\xi_1\D_{1,0}$ is precisely the collective
spin-$z$ operator $S_\z$ that appears in
Ref.~\cite{he2019engineering}.  Furthermore, drive operators $D_{LM}$
vanish when $L\ge n$, so in the case of SU(2) the operator $\D_{2,0}$
has no contribution to the effective Hamiltonians in
\eqref{eq:H_eff_1_phi} and \eqref{eq:H_eff_2_phi}.

@@

Here we derive the effective Hamiltonian induced on the fully
symmetric manifold $\M_\FS$ by the single-particle SOC Hamiltonian
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -t \sum_{q,L} w_L\p{q} A_L^{(\phi)} D_{L,0}^{(q)},
  &&
  w_L\p{q} \equiv
  \begin{cases}
    \cos q & L~\t{even} \\
    \sin q & L~\t{odd}
  \end{cases},
\end{align}
where $A_L^{(\phi)}$ is essentially a sum over Clebsch-Gordan
coefficients, defined in \eqref{eq:A_L_phi}, and scales as
$A_L^{(\phi)} \sim \phi^L + O\p{\phi^{L+2}}$ when $\abs{\phi}\ll1$.
For brevity, we work in units with the lattice spacing $a=1$ and
implicitly assume the limit $\abs{\phi}\ll1$ throughout this section.

In order to simplify the calculation of effective Hamiltonians, we
write the SOC Hamiltonian in the form
\begin{align}
  H_{\t{lat}}^{(\phi)} = \sum_{q,L} v_L^{(q)} D_{L,0}^{(q)},
  &&
  v_L^{(q)} \equiv - t w_L\p{q} A_L^{(\phi)},
\end{align}
and decompose it into its uniform and inhomogeneous parts as
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = \sum_L v_L \D_{L,0} + \sum_{q,L} z_L^{(q)} D_{L,0}^{(q)},
\end{align}
where
\begin{align}
  v_L \equiv \EE_q\sp{v_L^{(q)}},
  &&
  z_L^{(q)} \equiv v_L^{(q)} - v_L,
  &&
  \EE_q\sp{X_q} \equiv \f1N \sum_q X_q.
\end{align}

@@

The effective Hamiltonian in induced on the fully symmetric manifold
$\M_\FS$ at first order is
\begin{align}
  H_{\t{eff}}^{(1)}
  = \P_0 H_{\t{lat}}^{(\phi)} \P_0
  = \sum_L v_L \P_0 \D_{L,0} \P_0
  + \sum_{q,L} z_L^{(q)} \P_0 D_{L,0}^{(q)} \P_0,
\end{align}
where $\P_0$ is a projector onto $\M_\FS$.  Permutational symmetry
within the fully symmetric manifold $\M_\FS$ allow us to freely change
the quasi-momentum index on $\P_0D_{L,0}^{(q)}\P_0$, and the
coefficients $z_L^{(q)}$ are mean-zero by construction, so
\begin{align}
  \sum_{q,L} z_L^{(q)} \P_0 D_{L,0}^{(q)} \P_0
  = \sum_q z_L^{(q)} \P_0 D_{L,0}^{(0)} \P_0
  = 0,
\end{align}
which implies that
\begin{align}
  H_{\t{eff}}^{(1)}
  = \sum_L v_L \D_{L,0}
  = -t \sum_L \EE_q\sp{w_L\p{q}} A_L^{(\phi)} \D_{L,0},
  \label{eq:H_eff_1}
\end{align}
where we implicitly restrict $H_{\t{eff}}^{(1)}$ to the fully
symmetric manifold $\M_\FS$.

@@

The effective Hamiltonian in induced on the fully symmetric manifold
$\M_\FS$ at second order is\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_{\t{eff}}^{(2)}
  = - \P_0 H_{\t{lat}}^{(\phi)} \E H_{\t{lat}}^{(\phi)} \P_0
  &&
  \E \equiv \sum_{\Delta>0} \f{\P_\Delta}{\Delta},
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta$ above
that of fully symmetric manifold $\M_\FS$.  The uniform part of the
SOC Hamiltonian $H_{\t{lat}}^{(\phi)}$ is permutationally symmetric,
and therefore does not couple the fully symmetric manifold $\M_\FS$ to
its orthogonal complement.  As a consequence, the uniform part of
$H_{\t{lat}}^{(\phi)}$ is annihilated between the projector $\P_0$
onto $\M_\FS$ and the projectors in $\E$ onto the orthogonal
complement of $\M_\FS$.  We can therefore keep only the inhomogeneous
parts of $H_{\t{lat}}^{(\phi)}$ and expand
\begin{align}
  H_{\t{eff}}^{(2)}
  = -\sum_{J,K} \P_0 \Z_J \E \Z_K \P_0,
  &&
  \Z_L \equiv \sum_{q,L} z_L^{(q)} D_{L,0},
  \label{eq:H_eff_2_start}
\end{align}
We now divide the task of further simplifying $H_{\t{eff}}^{(2)}$ into
two steps:
\begin{enumerate*}
\item eliminating the excited-state projectors in $\E$, and
\item projecting the remaining product of operators in
  \eqref{eq:H_eff_2_start} onto the fully symmetric manifold.
\end{enumerate*}

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Eliminating excited-state projectors}

In order to eliminate the excited-state projectors in $\E$, we observe
that the action of $\Z_L$ on any fully symmetric state
$\ket{\psi_{\FS}}\in\M_\FS$ yields an eigenvector of the interaction
Hamiltonian $H_{\t{int}}=-u\sum_{p<q}\Pi_{pq}$.  To substantiate this
claim, we expand
\begin{align}
  \sum_{p<q} \Pi_{pq} \Z_L \ket{\psi_\FS}
  = \sum_{\substack{p<q\\k}}
  z_L^{(k)} \Pi_{pq} D_{L,0}^{(k)} \ket{\psi_\FS}
\end{align}
If $k\notin\set{p,q}$, then the permutation operator $\Pi_{pq}$
commutes with $D_{L,0}^{(k)}$, and in turn acts trivially on the
permutationally-symmetric state $\ket{\psi_\FS}$, giving us
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  z_L^{(k)} \Pi_{pq} D_{L,0}^{(k)} \ket{\psi_\FS}
  = \sum_k \sum_{\substack{p<q\\p,q\ne k}}
  z_L^{(k)} D_{L,0}^{(k)} \ket{\psi_\FS}
  = { N-1 \choose 2 } \Z_L \ket{\psi_\FS}.
  \label{eq:Z_eig_notin}
\end{align}
If $k\in\set{p,q}$, then we can resolve the identity operator
$\1=\Pi_{pq}\Pi_{pq}$ to expand
\begin{multline}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  z_L^{(k)} \Pi_{pq} D_{L,0}^{(k)} \ket{\psi_\FS}
  = \f12 \sum_{\substack{p\ne q\\k\in\set{p,q}}}
  z_L^{(k)} \Pi_{pq} D_{L,0}^{(k)} \Pi_{pq} \Pi_{pq} \ket{\psi_\FS}
  = \sum_q \sum_{k\ne q} z_L^{(k)} D_{L,0}^{(q)} \ket{\psi_\FS},
\end{multline}
where, using the fact that the coefficients $z_L^{(k)}$ are mean-zero,
\begin{align}
  \sum_{k\ne q} z_L^{(k)}
  = \sum_k z_L^{(k)} - z_L^{(q)}
  = - z_L^{(q)},
\end{align}
so
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  z_L^{(k)} \Pi_{pq} D_{L,0}^{(k)} \ket{\psi_\FS}
  = - \sum_q z_L^{(q)} D_{L,0}^{(q)} \ket{\psi_\FS}
  = - \Z_L \ket{\psi_\FS}.
\end{align}
In total,
\begin{align}
  \sum_{p<q} \Pi_{pq} \Z_L \ket{\psi_\FS}
  = \sp{{N-1\choose 2}-1} \Z_L \ket{\psi_\FS}.
\end{align}
The image of any fully symmetric state $\ket{\psi_\FS}\in\M_\FS$ under
the action of $\Z_L$ is thus proportional to an eigenstate of
$H_{\t{int}}$ with interaction energy $-u\sp{{ N-1 \choose 2 } - 1}$,
which is $Nu$ above the interaction energy $-u{N\choose2}$ of states
in the fully symmetric manifold, so
\begin{align}
  \E \Z_L \P_0
  = \sum_{\Delta>0} \f1{\Delta} \P_\Delta \Z_L \P_0
  =  \f1{Nu} \Z_L \P_0
  = \f1{fU} \Z_L \P_0,
  \label{eq:proj_elim}
\end{align}
where $f$ is the filling fraction of lattice sites and $U$ is the
two-body on-site interaction energy.  Our effective Hamiltonian thus
far is
\begin{align}
  H_{\t{eff}}^{(2)}
  = - \f1{fU} \sum_{J,K} \P_0 \Z_J \Z_K \P_0.
  \label{eq:H_eff_2_mid}
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Projecting onto the fully symmetric manifold}
